# Introduction

This project is based on the "Diabetes.csv" dataset, which contains multiple physiological indicators related to diabetes.

Our study has two main objectives:

1\. We use **binary logistic regression** to explore the relationships between different variables and diabetes, and to identify significant risk factors.

2\. We build a **decision tree** model to evaluate the effectiveness of diabetes prediction.

However, since the dataset contains missing values, we applied four imputation methods— **KNN, linear interpolation, missForest, and MICE,** to preprocess the data, and then compared whether the conclusions for the two research questions remain consistent under different imputation approaches.

# Data pre-processing and visualization

## Install and load necessary libraries.

```{r}
#install.packages("naniar")
#install.packages("VIM")
#install.packages("tidymodels")
#install.packages("rpart)
#install.packages("furrr", type = "binary")
#install.packages("rsample")
options(warn = -1)
set.seed(123)
library(rsample)
library(ggplot2)
library(naniar)
library(VIM)
library(simputation)
library(tidymodels)
library(rpart)
library(dplyr)
library(boot)
library(randomForest)
library(yardstick)
```

## Load the dataset and check basic information.

```{r}
# load the dataset
filepath = '/Users/caiqingming/Desktop/Diabetes.csv'
df <- read.csv(filepath, header = TRUE)
# replace all the "NA" values of the "Pregnancies" variable with zeros
df$Pregnancies[is.na(df$Pregnancies)] <- 0
# convert "Diabetes" variable into factor
df$Diabetes <- factor(df$Diabetes)
# check dimensions
dim(df)
# print first six rows
head(df)
```

## Check and visualize missing values and proportions.

```{r}
# count missing values for each variable
cat("=== Missing value count per column ===\n")
df %>%
  is.na() %>%
  colSums()
# calculate missing percentage
cat("=== Missing value percentage per column ===\n")
missing_percent <- colMeans(is.na(df)) * 100
print(round(missing_percent, 2))
```

```{r}
# visualize missing patterns
gg_miss_var(df)
```

```{r}
# visualize missing patterns
df %>% aggr(combined = TRUE ,numbers = TRUE, cex.axis = 0.75, cex.numbers = 0.7, oma = c(10, 5, 5, 2))
```

# Identify missing data mechanisms

**Task1**: test if the percentage of missing values in one variable differs for people with or without diabetes.

## Visualize missing values vs. Diabetes.

```{r}
# check missing in Insulin vs Diabetes
df %>% select(Diabetes, Insulin) %>% spineMiss()

# check missing in SkinThickness vs Diabetes
df %>% select(Diabetes, SkinThickness) %>% spineMiss()

# check missing in BloodPressure vs Diabetes
df %>% select(Diabetes, BloodPressure) %>% spineMiss()

# check missing in Glucose vs Diabetes
df %>% select(Diabetes, Glucose) %>% spineMiss()

# check missing in BMI vs Diabetes
df %>% select(Diabetes, BMI) %>% spineMiss()
```

## Perform Chi-square tests on missing values vs. Diabetes.

```{r}
# create dummy variables for missing values
df <- df %>% 
  mutate(
    missing_insulin = is.na(Insulin),
    missing_skinthickness = is.na(SkinThickness),
    missing_bloodpressure = is.na(BloodPressure),
    missing_bmi = is.na(BMI),
    missing_glucose = is.na(Glucose)
  )
```

### Insulin missing vs. Diabetes

```{r}
table_insulin <- table(df$missing_insulin, df$Diabetes)
cat("=== Contingency Table: Insulin missing vs Diabetes ===\n")
print(table_insulin)

# Chi-square test
cat("\n=== Chi-square Test ===\n")
chisq.test(table_insulin)
```

### SkinThickness missing vs. Diabetes

```{r}
table_skin <- table(df$missing_skinthickness, df$Diabetes)
cat("\n=== Contingency Table: SkinThickness missing vs Diabetes ===\n")
print(table_skin)

cat("\n=== Chi-square Test ===\n")
chisq.test(table_skin)
```

### BloodPressure missing vs. Diabetes

```{r}
table_bp <- table(df$missing_bloodpressure, df$Diabetes)
cat("\n=== Contingency Table: BloodPressure missing vs Diabetes ===\n")
print(table_bp)

cat("\n=== Chi-square Test ===\n")
chisq.test(table_bp)
```

### BMI missing vs. Diabetes

```{r}
table_bmi <- table(df$missing_bmi, df$Diabetes)
cat("\n=== Contingency Table: BMI missing vs Diabetes ===\n")
print(table_bmi)

cat("\n=== Chi-square Test ===\n")
chisq.test(table_bmi)
```

From results of Chi-square tests, we conclude that there is no significant relationship between missing values and Diabetes.

**Task2**: test if the percentage of missing values in one variable differs for different values of another continuous variable?

## Visualize missing values vs. continuous variables.

```{r}
# SkinThickness missing vs Age
marginplot(df[, c("Age", "SkinThickness")], 
           main = "Age vs SkinThickness Missing Pattern")

# SkinThickness missing vs BMI
marginplot(df[, c("BMI", "SkinThickness")], 
           main = "BMI vs SkinThickness Missing Pattern")

# Insulin missing vs Age
marginplot(df[, c("Age", "Insulin")], 
           main = "Age vs Insulin Missing Pattern")

# Insulin missing vs BMI
marginplot(df[, c("BMI", "Insulin")], 
           main = "BMI vs Insulin Missing Pattern")

# BMI missing vs Glucose
marginplot(df[, c("Glucose", "BMI")], 
           main = "Glucose vs BMI Missing Pattern")
```

## Perform t-test to verify missing values vs. continuous variables.

```{r}
# SkinThickness missing vs Age
missing_skin_age <- df %>% filter(missing_skinthickness == TRUE) %>% pull(Age)
complete_skin_age <- df %>% filter(missing_skinthickness == FALSE) %>% pull(Age)
t.test(missing_skin_age, complete_skin_age)

# SkinThickness missing vs BMI
missing_skin_bmi <- df %>% filter(missing_skinthickness == TRUE) %>% pull(BMI)
complete_skin_bmi <- df %>% filter(missing_skinthickness == FALSE) %>% pull(BMI)
t.test(missing_skin_bmi, complete_skin_bmi)

# Insulin missing vs Age
missing_insulin_age <- df %>% filter(missing_insulin == TRUE) %>% pull(Age)
complete_insulin_age <- df %>% filter(missing_insulin == FALSE) %>% pull(Age)
t.test(missing_insulin_age, complete_insulin_age)

# Insulin missing vs BMI
missing_insulin_bmi <- df %>% filter(missing_insulin == TRUE) %>% pull(BMI)
complete_insulin_bmi <- df %>% filter(missing_insulin == FALSE) %>% pull(BMI)
t.test(missing_insulin_bmi, complete_insulin_bmi)

# BMI missing vs Glucose
missing_bmi_glucose <- df %>% filter(missing_bmi == TRUE) %>% pull(Glucose)
complete_bmi_glucose <- df %>% filter(missing_bmi == FALSE) %>% pull(Glucose)
t.test(missing_bmi_glucose, complete_bmi_glucose)
```

For brevity, we only show tests with statistically significant results. All five tests show statistically significant differences (p \< 0.05).

**Result**:

(1) SkinThickness missing group has higher Age and lower BMI

(2) Insulin missing group has higher Age and lower BMI

(3) BMI missing group has lower Glucose

Since missing values are related to observed variables (Age, BMI, Glucose), the missing data mechanisms for **SkinThickness, Insulin and BMI** are **MAR** (Missing At Random).

To supplement the variable-specific t-tests, we performed **Little's MCAR test** to assess the global missing data pattern.

```{r}
df_original <- read.csv(filepath, header = TRUE)
mcar_test(df_original)
```

**Conclusion**: The result (p \< 0.05) rejects the MCAR hypothesis, confirming **a Missing at Random (MAR) mechanism** and justifying the use of imputation methods below.

# Perform different imputation methods.

## Perform KNN imputations

### Normalize numeric columns

```{r}
# identify numeric columns
num_cols <- names(df)[sapply(df, is.numeric)]

# normalize (min-max scaling)
mins <- sapply(df[num_cols], min, na.rm = TRUE)
maxs <- sapply(df[num_cols], max, na.rm = TRUE)

df_scaled <- df
for (col in num_cols) {
  df_scaled[[col]] <- (df_scaled[[col]] - mins[col]) / (maxs[col] - mins[col])
}
```

### Hyperparameter tuning for K in KNN

```{r}
# identify columns with missing values
target_vars <- names(which(colSums(is.na(df_scaled)) > 0))

# extract complete cases
df_complete <- df_scaled[complete.cases(df_scaled), ]

# define parameter grid
k_grid <- seq(1, 141, by = 6)
n_repeats <- 10

# store RMSE for each K
rmse_sum <- rep(0, length(k_grid))

# repeated masking and evaluation
for (rep in 1:n_repeats) {
  set.seed(123 + rep)
  
  # mask 10% of complete data
  mask_idx <- sample(1:nrow(df_complete), nrow(df_complete) * 0.1)
  df_masked <- df_complete
  true_vals <- as.matrix(df_complete[mask_idx, target_vars])
  df_masked[mask_idx, target_vars] <- NA
  
  # evaluate each K
  for (i in seq_along(k_grid)) {
    imp_tmp <- kNN(df_masked, k = k_grid[i], imp_var = FALSE)
    pred_vals <- as.matrix(imp_tmp[mask_idx, target_vars])
    rmse_sum[i] <- rmse_sum[i] + sqrt(mean((pred_vals - true_vals) ^ 2, na.rm = TRUE))
  }
}

# calculate average RMSE across repeats
mean_rmse <- rmse_sum / n_repeats

best_k <- k_grid[which.min(mean_rmse)]
print(best_k)

# Visualize results
tune_df <- data.frame(K = k_grid, RMSE = mean_rmse)

ggplot(tune_df, aes(x = K, y = RMSE)) +
  geom_line(color = "steelblue") +
  geom_point(color = "red") +
  geom_vline(xintercept = best_k, linetype = "dashed") +
  labs(x = "K", y = "RMSE", title = paste("Best K =", best_k))
```

**Conclusion**: According to the tuning plot, we observe two critical points:

(1) **The elbow point (K=7)** where the error drop stabilizes. We choose this to preserve local data structure.

(2) **The global minimum (K=103)** where the RMSE is lowest due to maximum smoothing.

We will compare the effects of choosing different values of K.

### Perform KNN imputations with the chosen K.

```{r}
# K = 7
# sort variables by number of missing values
vars_by_NAs <- df_scaled %>% 
  is.na() %>% 
  colSums() %>% 
  sort(decreasing = FALSE) %>% 
  names()
# perform imputations
df_scaled_imp_1 <- df_scaled %>% 
  select(all_of(vars_by_NAs)) %>% 
  kNN(k = 7, imp_var = TRUE)

# reverse to original scale
df_imp_1 <- df_scaled_imp_1
for (col in num_cols) {
  df_imp_1[[col]] <- df_scaled_imp_1[[col]] * (maxs[col] - mins[col]) + mins[col]
}

# visualize imputations
df_imp_1 %>% 
  select(Glucose, Insulin, Insulin_imp) %>% 
  marginplot(delimiter = "_imp", 
             main = "KNN Imputation Results (K=7) - Insulin",
             col = c("skyblue", "orange"))

# check is there are no missing values
colSums(is.na(df_imp_1))
# remove extra columns
df_original <- read.csv(filepath, header = TRUE)
df_imp_1 <- df_imp_1[names(df_original)]
```

```{r}
# K = 103
# sort variables by number of missing values
vars_by_NAs <- df_scaled %>% 
  is.na() %>% 
  colSums() %>% 
  sort(decreasing = FALSE) %>% 
  names()
# perform imputations
df_scaled_imp_2 <- df_scaled %>% 
  select(all_of(vars_by_NAs)) %>% 
  kNN(k = 103, imp_var = TRUE)

# reverse to original scale
df_imp_2 <- df_scaled_imp_2
for (col in num_cols) {
  df_imp_2[[col]] <- df_scaled_imp_2[[col]] * (maxs[col] - mins[col]) + mins[col]
}

# visualize imputations
df_imp_2 %>% 
  select(Glucose, Insulin, Insulin_imp) %>% 
  marginplot(delimiter = "_imp", 
             main = "KNN Imputation Results (K=103) - Insulin",
             col = c("skyblue", "orange"))

# check is there are no missing values
colSums(is.na(df_imp_2))
# remove extra columns
df_original <- read.csv(filepath, header = TRUE)
df_imp_2 <- df_imp_2[names(df_original)]
```

### Apply logistic regression to identify diabetes risk factors.

#### When K = 7

```{r}
# K = 7
# fit logistic regression
model1 <- glm(Diabetes ~ ., 
             data = df_imp_1, 
             family = binomial)

summary(model1)

# extract results
result1 <- data.frame(
  Estimate = coef(model1),
  OR = exp(coef(model1)),
  p_value = summary(model1)$coefficients[,4]
)

print(result1)
```

#### When K = 103

```{r}
# K = 103
# fit logistic regression
model2 <- glm(Diabetes ~ ., 
             data = df_imp_2, 
             family = binomial)

summary(model2)

# extract results
result2 <- data.frame(
  Estimate = coef(model2),
  OR = exp(coef(model2)),
  p_value = summary(model2)$coefficients[,4]
)

print(result2)
```

**Comparison**:

(1) The most significant shift between K=7 and K=103 is in **Insulin**, where p-value changes from **0.091** to **0.003**, making result statistically significant.

(2) Both models show that **Pregnancies, Glucose, BMI and DiabetesPedigreeFunction** have significant positive relationships with **Diabetes**.

### Use decision tree to forecast diabetes.

#### When K = 7

```{r}

df_imp_1$Diabetes <- factor(df_imp_1$Diabetes)

# split into training and testing sets
set.seed(123)
data_split_1 <- initial_split(df_imp_1, prop = 0.75, strata = Diabetes)
train_data_1 <- training(data_split_1)
test_data_1 <- testing(data_split_1)

# create decision tree specification
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

# fit the model to the training data
tree_fit <- tree_spec %>%
  fit(Diabetes ~ ., data = train_data_1)

# make predictions on the testing data
predictions <- tree_fit %>%
  predict(test_data_1) %>%
  pull(.pred_class)

# calculate evaluation results
results_1 <- data.frame(
  truth = test_data_1$Diabetes,
  pred = predictions
)

# calculate metrics
metric_set(accuracy, f_meas, yardstick::precision, yardstick::recall)(
  results_1,
  truth = truth,
  estimate = pred,
  event_level = "second"
)

cat("\n=== Confusion Matrix ===\n")
cm_1 <- conf_mat(results_1, truth = truth, estimate = pred)
cm_1

```

#### When K = 103

```{r}

df_imp_2$Diabetes <- factor(df_imp_2$Diabetes)

# split into training and testing sets
set.seed(123)
data_split_2 <- initial_split(df_imp_2, prop = 0.75, strata = Diabetes)
train_data_2 <- training(data_split_2)
test_data_2 <- testing(data_split_2)

# create decision tree specification
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

# fit the model to the training data
tree_fit <- tree_spec %>%
  fit(Diabetes ~ ., data = train_data_2)

# make predictions on the testing data
predictions <- tree_fit %>%
  predict(test_data_2) %>%
  pull(.pred_class)

# calculate evaluation results
results_2 <- data.frame(
  truth = test_data_2$Diabetes,
  pred = predictions
)

# calculate metrics
metric_set(accuracy, f_meas, yardstick::precision, yardstick::recall)(
  results_2,
  truth = truth,
  estimate = pred,
  event_level = "second"
)

cat("\n=== Confusion Matrix ===\n")
cm_2 <- conf_mat(results_2, truth = truth, estimate = pred)
cm_2

```

**Comparison**: **K=103** produces higher predictive performance than **K=7**. While **K=7** is statistically more realistic, the smoothed data from K=103, is more effective for maximizing predictive power by filtering out local noise.

## Perform interative Linear Regression imputations

```{r}
df <- read.csv(filepath, header = TRUE)
df$Diabetes <- as.factor(df$Diabetes)

df_imp_LR <- hotdeck(df)

# Store the locations of missing values in the original data
missing_preg <- is.na(df$Pregnancies)
missing_glu  <- is.na(df$Glucose)
missing_bp   <- is.na(df$BloodPressure)
missing_skin <- is.na(df$SkinThickness)
missing_ins  <- is.na(df$Insulin)
missing_bmi  <- is.na(df$BMI)

# Iterative imputation
n_iter <- 5

for (i in 1:n_iter) {
  cat("Iteration:", i, "\n")
  
  # Impute Pregnancies
  df_imp_LR$Pregnancies[missing_preg] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    Pregnancies ~ Glucose + BloodPressure + SkinThickness + Insulin +
      BMI + DiabetesPedigreeFunction + Age + Diabetes
  )
  
  # Impute Glucose
  df_imp_LR$Glucose[missing_glu] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    Glucose ~ Pregnancies + BloodPressure + SkinThickness + Insulin +
      BMI + DiabetesPedigreeFunction + Age + Diabetes
  )
  
  # Impute BloodPressure
  df_imp_LR$BloodPressure[missing_bp] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    BloodPressure ~ Pregnancies + Glucose + SkinThickness + Insulin +
      BMI + DiabetesPedigreeFunction + Age + Diabetes
  )
  
  # Impute SkinThickness
  df_imp_LR$SkinThickness[missing_skin] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    SkinThickness ~ Pregnancies + Glucose + BloodPressure + Insulin +
      BMI + DiabetesPedigreeFunction + Age + Diabetes
  )
  
  # Impute Insulin
  df_imp_LR$Insulin[missing_ins] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    Insulin ~ Pregnancies + Glucose + BloodPressure + SkinThickness +
      BMI + DiabetesPedigreeFunction + Age + Diabetes
  )
  
  # Impute BMI
  df_imp_LR$BMI[missing_bmi] <- NA
  df_imp_LR <- impute_lm(
    df_imp_LR,
    BMI ~ Pregnancies + Glucose + BloodPressure + SkinThickness +
      Insulin + DiabetesPedigreeFunction + Age + Diabetes
  )
}

# Preview the imputed dataset
head(df_imp_LR)

```

### Apply logistic regression to identify diabetes risk factors.

```{r}
model <- glm(Diabetes ~ ., 
             data = df_imp_LR, 
             family = binomial)
summary(model)
coef_table <- summary(model)$coefficients

result <- data.frame(
  Estimate = coef_table[, "Estimate"],
  OR = exp(coef_table[, "Estimate"]),
  p_value = coef_table[, "Pr(>|z|)"]
)

print(result)
```

### Use decision tree to forecast diabetes.

```{r}
set.seed(123)
data_split_LR <- initial_split(df_imp_LR, prop = 0.75)
train_data_LR <- training(data_split_LR)
test_data_LR  <- testing(data_split_LR)
```

```{r}
## training
tree_spec <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

tree_fit <- tree_spec %>%
  fit(Diabetes ~ ., data = train_data_LR)
```

```{r}
## evaluation
rm(accuracy, f_meas, precision, recall)
my_metrics <- metric_set(accuracy, f_meas, precision, recall)

model_performance_LR <- tree_fit %>%
  predict(test_data_LR, type = "class") %>%
  bind_cols(test_data_LR) %>%
  mutate(
    Diabetes = factor(Diabetes, levels = c("no", "yes")),
    .pred_class = factor(.pred_class, levels = c("no", "yes"))
  ) %>%
  my_metrics(truth = Diabetes, estimate = .pred_class, event_level = "second")

print(model_performance_LR)
```
